Polynomial Chaos for the Estimation of Manning's Friction

### 2013-12-17 paper/intro.tex
added two paragraphs to the intro but needs some cleaning

Previous work looking into uncertainty in the input for tsunami simulations often has focused on the earthquake source but mainly as a way of constraining the various proposed earthquake scenarios only after the event (see for example \cite{MacInnes:2013cr}).  In this work we concentrate on the uncertainty in the friction parameterization used in tsunami simulations.  Changes to the friction parameterization has been shown in the past to lead to significant changes to inundation levels \cite{Myers:2001el,Jakeman:2010hk} but these previous studies were limited in the scope of the changes allowed.

In this work we have used the widely used Manning's-n law to provide a relationship between the friction coefficient and the roughness parameter $n$.  The parameter $n$ is also allowed to vary in space determined by the initial water depth to study effects of deep-water, near-shore, and on-shore sensitivities.  To carry out the uncertainty quantification we have employed polynomial chaos expansions as used in \cite{sraj:2013a,sraj:2013b}.  Here the quantity of interest is taken to be the surface elevation $\eta$ and will compared to available NOAA Deep-ocean Assessment and Reporting of Tsunamis (DART) buoys for the \tohoku tsunami.  The forward model runs are performed using \geoclaw following the setup and using one of the source models explored in \cite{MacInnes:2013cr}.

\begin{itemize}
    \item Possibly mention this \url{http://arxiv.org/pdf/1203.6297.pdf}?  Talks about creating a surrogate model for land-slide generated tsunamis.
\end{itemize}

>>>>

Previous work looking into uncertainty in the input for tsunami simulations often has focused on the earthquake source but mainly as a way of constraining the various proposed earthquake scenarios only after the event (see for example \cite{MacInnes:2013cr}).  In this work we concentrate on the uncertainty in the friction parameterization used in tsunami simulations.  Changes to the friction parameterization has been shown in the past to lead to significant changes to inundation levels \cite{Myers:2001el,Jakeman:2010hk} but these previous studies were limited in the scope of the changes allowed.  Recently, Sarri \emph{et al.} built a statistical surrogate of a landslide-generated tsunami computer model by using an emulator~\cite{Sarri2012}. The implemented emulator was the Gaussian process
that requires using a combination of prior knowledge about the simulator and
appropriate choices of functions and parameters. Their work was presented as a proof-of-concept case study
where they performed basic statistical and sensitivity analysis

In this work, we take advantage of the recent developments in Uncertainty Quantification 
(UQ) methods, which make it feasible to probe the sensitivity of complex and realistic tsunami models 
to uncertain parameters, and to assess the amount of information that can
be gained from specific measurements and observations. In particular, 
we implement an inverse modeling approach towards the objective stated above.  
Specifically, we will rely on Polynomial Chaos (PC) expansions to construct a 
faithful surrogate of the response of the large and complex geophysical 
model simulations.  The availability of the surrogate enables us to efficiently 
implement a Bayesian inference formalism to the inverse problem, namely 
because it dramatically reduces the cost of sampling the posterior distribution.

PC expansions have been developed and applied in the engineering community to quantify uncertainties
in numerical simulations; their principal use is in propagating input
uncertainties through large, complex and non-linear models to compute the
ensuing output uncertainties. Several researchers have been 
investigating the applicability of PC expansions for UQ in oceanic simulations
\citep{thacker2012,ashwanth2010,Alexanderian2012}. Recently,
PC combined with Bayesian inference was used to estimate drag coefficient
during Typhoon Fanapi 2010. The inference was done using
adjoint method and MCMC \cite{sraj:2013a,sraj:2013b}.

In this work, we follow a similar procedure as in \cite{sraj:2013a,sraj:2013b}
to estimate Manning N friction. We use the widely used Manning's-n law to provide a relationship between the friction coefficient and the roughness parameter $n$.  The parameter $n$ is also allowed to vary in space determined by the initial water depth to study effects of deep-water, near-shore, and on-shore sensitivities.  To carry out the uncertainty quantification we have employed polynomial chaos expansions.  Here the quantity of interest is taken to be the surface elevation $\eta$ and will compared to available NOAA Deep-ocean Assessment and Reporting of Tsunamis (DART) buoys for the \tohoku tsunami.  The forward model runs are performed using \geoclaw following the setup and using one of the source models explored in \cite{MacInnes:2013cr}.
